High geomagnetic field intensity recorded by anorthosite xenoliths requires a vigorous late Mesoproterozoic geodynamo

### 2021-12-20 PNAS_manuscript/PNAS_manuscript.tex

% Please add a significance statement to explain the relevance of your work
\significancestatement{Acquiring high-fidelity observational records of ancient magnetic field intensity from remanent magnetization of rocks is crucial for constraining the evolution history of Earth's core. However, robust estimates of ancient field strengths are often difficult to recover due to alterations or non-ideal rock magnetic behaviors. In this study, we use plagioclase cumulates that formed in the deeper crust and were brought to the near surface by magma where they cooled and acquired thermal remanent magnetizations. These plagioclase from the North American Midcontinent Rift have experienced minimal post-formation alterations and record a present-day-like virtual dipole moment 1.1 billion years ago. The results indicate the existence of strong power sources for Earth's dynamo during the late Mesoproterozoic. }

@@

\begin{abstract}
Obtaining estimates of Earth's magnetic field strength in deep time is complicated by non-ideal rock magnetic behavior in many igneous rocks. In this study, we target \textit{ca.} 1092 Ma anorthosite xenoliths that cooled and acquired their magnetization within Beaver River diabase intrusions of the North American Midcontinent Rift. In contrast to the diabase which has non-ideal paleointensity behaviors and fails to provide reliable paleointensity estimates, the anorthosite xenoliths are unusually high fidelity recorders yielding high-quality, single-slope paleointensity results that are consistent at specimen and site levels. An average value of $\sim$89 ZAm$^2$ for the virtual dipole moment from the anorthosites is higher than that of the dipole component of Earth's magnetic field today. Such a strong late Mesoproterozoic dipole recorded by the anorthosites requires the existence of a vigorous geodynamo at the time. Together with previous paleointensity data from other Midcontinent Rift rocks, our new results indicate that a dynamo with strong power sources persisted through $\sim$25 Myr at \textit{ca.} 1.1 Ga. These observations are inconsistent with there being a progressive monotonic decay of Earth's dynamo strength through the Proterozoic Eon.
\end{abstract}

@@

Paleomagnetic records from ancient rocks are one of the few types of observational data that have the potential to record the thermal evolution of Earth’s core. Evidence for a persistent magnetic field through the Proterozoic, for example, necessitates the existence of plate tectonics that sustained core-mantle boundary heat flow. However, strikingly low estimates of geomagnetic field strengths have been obtained in the Ediacaran Period (ref. \citealp{Bono2019a}; \textit{ca.} 565 Ma), potentially indicating an unusual period of geodynamo regime at the time.

@@

Determinations of ancient magnetic field strengths rely on igneous rocks that acquire thermal remanent magnetizations as they cool in Earth's magnetic field. These magnetizations need to be unmodified by subsequent heating or chemical alterations in order to be able to faithfully record ancient geomagnetic field record through time. Intraplate magmatic events are therefore an important target as they can be well-preserved within continental interiors at distance from mountain-building events along continental margins. However, magmatic events associated with large igneous provinces are typically of geologically short duration with the bulk of magmatic products emplaced within $<$1 Myr. The Midcontinent Rift is an exception as magmatism lasted $\sim$25 Myr from 1109 Ma to 1084 Ma. Additionally, extension ceased in the Midcontinent Rift prior to lithosphere separation, preserving volcanics, intrusives, and sediments of the rift within the continental interior. As a result, rocks of the rift have simple paleomagnetic behavior and form a central record of Mesoproterozoic paleogeography \cite{Swanson-Hysell2021c}. This simple magnetization and the long duration of magnetization holds the potential for a well-preserved extended window into the intensity of Earth's magnetic field from 1109 to 1084 Ma.

Despite the exquisite preservation of the rocks, non-ideal paleointensity behaviors and overall low success rates have often led to high failure rates and challenged the interpretation of many previous paleointensity results from the Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a}. One typical failure of paleointensity experiment is the double-slope behavior, as two distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope: a higher paleointensity estimate from the low-temperature portion and a lower paleointensity estimate from the high-temperature portion. Ref \citealp{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas ref. \citealp{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Results with such non-ideal behavior were rejected by ref. \citealp{Sprain2018a} who applied stricter paleointensity selection criteria.

In this study, we target a unique rock type - high-purity anorthosite xenoliths. These anorthosite xenoliths were brought to the near surface with magma that formed sills of the Beaver River diabase. They cooled and acquired their magnetizations in conjunction with the host diabase. Paleointensity experiments on the anorthosite xenoliths have a very high success rate, yielding consistent specimen- and site-level paleointensity results. Rock magnetic data reveal that the anorthosite xenoliths have low anisotropy of thermal remanent magnetization (TRM) and can acquire TRM linearly. Magnetic imaging shows the anorthosite specimens have dominant magnetic carriers interstitial to and within plagioclase crystals and they do not display strong preferred orientations. Our single-slope, high quality paleointensity data add to the reliable paleointensity records during the $\sim$25 Myr of Midcontinent Rift magmatic activity. Our new data from the Midcontinent Rift intrusive rocks support previous data from the volcanics that there was a strong late Mesoproterozoic geodynamo.

@@

While magmatic activity within the Midcontinent Rift was protracted, there were volcanism and intrusions characterized by particularly rapid and voluminous emplacement. The emplacement of the \textit{ca.} 1092 Ma Beaver Bay Complex in northern Minnesota punctuates one such period of magmatism during the Midcontinent Rift main stage activity. The magma that formed the 1091.7 $\pm$ 0.2 Ma Beaver River diabase dikes and sills of the Beaver Bay Complex transported numerous anorthosite xenoliths that have short-axis diameters up to 180 meters via wide conduits \cite{Boerboom2004a, Boerboom2006b}. The anorthosite xenoliths are plagioclase cumulates that formed comagmatically with the host diabase in the lower crust - an interpretation confirmed by U-Pb zircon geochronology \cite{Zhang2021b}.

@@

Thermal modeling results and paleomagnetic directional data show the anorthosite xenoliths to have acquired thermal remanent magnetizations while cooing with the Beaver River diabase \cite{Zhang2021b}. Step-wise thermal demagnetizations show the anorthosite xenoliths have dominantly single-component magnetization that often unblock sharply within temperature ranges between 500\textdegree C and 580\textdegree C.

@@

Petrographic imaging shows the dominantly monomineralic anorthosite xenoliths of the Beaver River diabase often develop granoblastic texture characterized by equigranular crystals with poor development of petrofabrics (Fig. \ref{fig:Petro_QDM}A). In addition, the plagioclase crystals of the anorthosite xenoliths are often free of large inclusions. Magnetic imaging using the quantum diamond microscope (QDM) shows that the dominant magnetic remanence carriers often exist within and sometimes interstitial to plagioclase crystals (Fig. \ref{fig:Petro_QDM}E).

These features of the Beaver River anorthosite xenoliths are distinct from other anorthosites such as the anorthosite xenoliths within the Duluth Complex Anorthositic Series rocks. The plagioclase of the Duluth Complex anorthosites typically develop interlocking textures that display strong igneous foliations (Fig. \ref{fig:Petro_QDM}B) and associated magnetic mineral fabrics characterized by having aligned titanomagnetite needles within plagioclase crystals (Fig. \ref{fig:Petro_QDM}D). Magnetic imaging confirms that the titanomagnetite needles within the plagioclase crystals are often strongly magnetic with magnetic fabrics oriented along the long axes of the needles (Fig. \ref{fig:Petro_QDM}F). Titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex plagioclase \cite{Miller1990a}.

The lack of symplecites and exsolved titanomagnetite needles within the Beaver River anorthosite xenoliths reflects that the Beaver River diabase magma had a different oxidation history than that of the Duluth Complex. The plagioclase and oxide mineral textures of the Beaver River anorthosites also distinguish them from plagioclase cumulates of other layered mafic complex where magnetic mineral fabrics associated with igneous foliation often occur \cite{Scofield1986a, Selkin2000a, Feinberg2006a}.

@@

Following IZZI-modified paleointensity experiments, 40 from a total of 86 anorthosite specimens and 0 out of a total of 69 diabase specimens passed our paleointensity result selection criteria of having (1) a maximum angular deviation (MAD; ref. \citealp{Kirschvink1980a}) of $<$20\textdegree; (2) scatter parameter ($\beta$; ref. \citealp{Coe1978a}) values of $<$15 per cent; (3) a deviation angle (DANG; ref. \citealp{Tauxe2004a}) of $<$5\textdegree; (4) fraction of remanence fitted for paleointensity estimate (FRAC; ref. \cite{Shaar2013a}) $>$0.6; (5) scatter statistic (SCAT; ref. \citealp{Shaar2013a}) = TRUE; (6) a maximum magnetic moment difference between adjacent zero-field steps (GAP-Max; ref. \citealp{Shaar2013a}) $<$ 0.25; (7) number of pTRM checks $>$ 2; (8) and number of measurements used for paleointensity determination $\geq$ 4. 7 anorthosite sites and no diabase sites with site-level paleointensity results pass our selections. Example NRM/TRM plots are shown in Fig. \ref{fig:IZZI_examples}. The cooling rate corrected absolute paleointensity estimates from all anorthosite sites have a weighted mean of 52.72 $\pm $5.72 $\mu$T. This corresponds to a calculated virtual dipole moment of $\sim$89 ZAm$^2$ \textit{ca.} 1092 Ma. Summary specimen absolute paleointensity estimates and site-level weighted mean paleointensity values are plotted in Fig. \ref{fig:PINT_cooling_corrected} (and provided in Table S1). All measurement-level paleointensity experiment data is available within the MagIC database (\url{}; UPDATE WHEN DOI IS GENERATED).

@@

Paleointensity experiment results from both lithologies show three general types of dominant specimen behaviors on NRM/TRM plots (Fig. \ref{fig:IZZI_examples}): dominantly single-slope, double-slope or sagging, and zigzagging. Typical specimen paleointensity behavior of the anorthosites display straight, single-slope NRM/TRM plots and the accepted fractions of temperature steps span over the origin-trending, primary remanence components (Fig. \ref{fig:IZZI_examples}). We accept all specimen- and site-level absolute paleointensity results from the anorthosite xenoliths that pass the selection. Diabase specimens failed the selection because of double-slope or sagging behavior (fail FRAC selection; see methods), poor pTRM checks, and zigzagging (fail SCAT, DANG selection; see methods) plots (Fig. \ref{fig:IZZI_examples}). A 20 mT alternating field treatment after in-field heating steps was applied to some specimens but it did not result in significant changes in paleointensity results from the anorthosite xenoliths and the diabase (Fig. \ref{fig:PINT_cooling_corrected}; Supporting Information).

In addition to estimating paleointensities by introducing a set of selection criteria to filter our experiment results, we apply an independent method of ref. \citealp{Cych2021a} which uses all experimental data regardless of their NRM/TRM plot statistics to perform bias corrected estimation of paleointensity (BiCEP). This BiCEP method is a Bayesian probabilistic method based on an assumption that paleointensity estimates from specimens that come from a same cooling unit are distributed around a true paleointensity value with the various deflections being expressed as the curvature parameter of the NRM/TRM plot \cite{Arai1963a, Paterson2011a}. The posterior paleointensity distributions for sites with high-quality specimen-level data are in agreement with site-level weighted-mean results (Fig. S3), consistent with the interpretation that the anorthosite xenoliths record a high geomagnetic field \textit{ca.} 1092 Ma.

Overall, the consistent specimen- and site-level behavior makes the paleointensity results from the anorthosites of high quality. The paleointensity quality index (Q$_{PI}$; ref. \citealp{Biggin2014a}) for the anorthosite sites are all above 4 (Table S2).

@@

Rock magnetic data support that the successful anorthosite specimens have dominant magnetic remanence carriers with magnetic properties similar to stochiometric, non-interacting, single domain magnetite, whereas anorthosite samples that failed the paleointensity result selection and all diabase samples have more pronounced populations of non-ideal carriers. Magnetic property measurement system (MPMS) data show both diabase and anorthosite have dominant magnetic mineralogy of magnetite as evidenced through the Verwey transitions (Fig. S1; ref. \citealp{Verwey1939a}). Anorthosite chips from sites that yield successful paleointensity results have Verwey transition temperatures near 120 Kelvin (K) as expected for stochiometric low-Ti magnetite \cite{Ozdemir1993a}. However, diabase and anorthosite specimens that did not pass our paleointensity selection typically have suppressed Verwey transition toward lower temperatures (Fig. S1), indicating that the specimens likely have a magnetic mineralogy with variable degrees of oxidation or relatively high Ti content \cite{Ozdemir1993a}.

Coercivity spectra from backfield experiments further support the interpretation that rocks that pass or fail the selection criteria have distinct magnetic mineralogy. A compilation of median destructive field (MDF) in Fig. \ref{fig:coercivity} based on our backfield experiments shows that specimens from anorthosites that pass paleointensity selections have a distinctly higher average MDF range than other anorthosites and diabase. Example single component fits for coercivity spectra \cite{Maxbauer2016a} show that anorthosites that yielded successful paleointensity results can have magnetic grain populations with peak coercivity around 80 mT (Fig. \ref{fig:coercivity}). In contrast, other anorthosite and diabase tend to have lower coercivities ($\sim$30 mT). This is consistent with the interpretation that they lack a dominance of single domain magnetite grains in their magnetic mineral assemblages.

\begin{figure}
\noindent\includegraphics[width=\linewidth]{coercivity.pdf}
\centering
\caption{\footnotesize{Top: Example coercivity unmixing plots of anorthosite and diabase specimens from sites that pass or fail our paleointensity selection criteria. Bottom: Box plots of median destructive field (MDF) values for all anorthosite and diabase specimens with single-component coercivity unmixing results. Both plots show that anorthosite specimens that pass our paleointensity selection criteria have distinctly higher MDF than other anorthosites that do not pass the selection, as well as all diabase specimens.}}
\label{fig:coercivity}
\end{figure}

\subsection*{TRM anisotropy and nonlinear acquisition check}
Significant remanence anisotropy has been documented to exist within anorthositic rocks that form in layered intrusive complexes \cite{Selkin2000a, Feinberg2006a}. Ref. \citealp{Selkin2000a} demonstrated that strong remanence anisotropy associated with the igneous foliation developed within an anorthosite sample from the Stillwater Complex can lead to significant overestimate or underestimate of paleointensity values depending on the relative orientations between the external applied field magnetic anisotropy axes of a specimen. To assess whether our paleointensity estimates are biased by remanence anisotropy, we calculated the gamma statistic, which is the angular difference between the last pTRM step of paleointensity experiment and the applied field direction. The results show that the anorthosite specimens used for paleointensity experiment have gamma values ranging from $\sim$0.9\textdegree  to $\sim$11.9\textdegree, with a median value of 4.2\textdegree (Table S1). These gamma values are similar to those of the Midcontinent Rift volcanics \cite{Sprain2018a}. Therefore, the Beaver River anorthosite bulk samples do not have a significant remanence anisotropy. Paleodirectional data from our anorthosite xenoliths further support that they have minimal remanence anisotropy as their site mean directions closely match those of the Beaver River diabase hosts \cite{Zhang2021b}. In addition, no strong magnetic fabric was observed through magnetic imaging (Fig. \ref{fig:Petro_QDM}).

We applied full TRMs to our anorthosite specimens that pass the paleointensity selection criteria with lab fields of 30, 50, 70, and 90 $\mu$T along specimen vertical axes. The results show that our anorthosite specimens did not saturate upon the applied fields and there is thus no need for correcting nonlinear thermal remanence acquisitions (Fig. S4).

@@

\subsection*{Cooling rate correction}
Paleointensity estimates from igneous intrusions can also be biased toward higher values due to slow cooling rates \cite{Halgedahl1980a}. It has been shown that large differences in cooling rates between acquisition of an NRM versus a laboratory pTRM can result in overestimated paleointensities for SD grains \cite{Dodson1980a, Halgedahl1980a, Nagy2021a}. From the thermal history model of ref. \citealp{Zhang2021b} we can estimate the duration which the diabase and anorthosite cooled from the Curie temperature of magnetite ($\sim$580\textdegree C) to the time when they fully block in their characteristic natural remanence magnetization ($\sim$500\textdegree C; ref. \citealp{Zhang2021b}). We find the cooling time to be on the order of $\sim$1 kyr, which corresponds to a cooling rate of $\sim 1.6\times10^{-9}$ $^\circ$C s$^{-1}$. In contrast, the lab cooling rate is much faster, at $\sim 1.3\times10^{-1}$ $^\circ$C s$^{-1}$, assuming that it takes 1000s in lab condition to cool through the same temperature interval. This significant cooling rate difference leads to a predicted $\sim$36\% overestimate of true ancient field following the model of ref. \citealp{Halgedahl1980a} (Fig. S2). This estimate on cooling rate effect is similar to that derived from the model of ref. \citealp{Nagy2021a}. We therefore correct our paleointensity results by a factor of 0.74. The cooling rate-corrected specimen paleointensity estimates together with specimen- and site-level weighted means are shown in Fig. \ref{fig:PINT_cooling_corrected}. Taking all accepted anorthosite paleointensity results, they give a site mean absolute paleointensity estimate of 52.72 $\pm$ 5.72 $\mu$T (Fig. \ref{fig:PINT_cooling_corrected}).

>>>>

% Please add a significance statement to explain the relevance of your work
\significancestatement{Acquiring high-fidelity observational records of ancient magnetic field intensity from remanent magnetization of rocks is crucial for constraining the evolution of Earth's core. However, robust estimates of ancient field strengths are often difficult to recover due to alterations or non-ideal rock magnetic behaviors. In this study, we use plagioclase cumulates that formed in the deep crust and were brought to the near surface by magma where they cooled and acquired thermal remanent magnetizations. These plagioclase from the North American Midcontinent Rift have experienced minimal post-formation alterations and record a present-day-like virtual dipole moment 1.1 billion years ago. The results indicate the existence of strong power sources for Earth's dynamo during the late Mesoproterozoic. }

@@

\begin{abstract}
Obtaining estimates of Earth's magnetic field strength in deep time is complicated by non-ideal rock magnetic behavior in many igneous rocks. In this study, we target \textit{ca.} 1092 Ma anorthosite xenoliths that cooled and acquired their magnetization within Beaver River diabase intrusions of the North American Midcontinent Rift. In contrast to the diabase which has non-ideal paleointensity behaviors and fails to provide reliable paleointensity estimates, the anorthosite xenoliths are unusually high-fidelity recorders yielding high-quality, single-slope paleointensity results that are consistent at specimen and site levels. An average value of $\sim$89 ZAm$^2$ for the virtual dipole moment from the anorthosites is higher than that of the dipole component of Earth's magnetic field today. Such a strong late Mesoproterozoic dipole recorded by the anorthosites requires the existence of a vigorous geodynamo at the time. Together with previous paleointensity data from other Midcontinent Rift rocks, our new results indicate that a dynamo with strong power sources persisted through $\sim$25 Myr at \textit{ca.} 1.1 Ga. These observations are inconsistent with there being a progressive monotonic decay of Earth's dynamo strength through the Proterozoic Eon.
\end{abstract}

@@

Paleomagnetic records from ancient rocks are one of the few types of observational data that have the potential to record the thermal evolution of Earth’s core. Evidence for a persistent magnetic field through the Proterozoic, for example, necessitates the existence of plate tectonics that sustained core-mantle boundary heat flow. However, strikingly low estimates of geomagnetic field strengths have been obtained in the Ediacaran Period \cite{Bono2019a, Thallner2021b}, potentially indicating an unusual period of core dynamo regime at the time.

@@

Determinations of ancient magnetic field strengths rely on igneous rocks that acquire thermal remanent magnetizations as they cool in Earth's magnetic field. These magnetizations need to be unmodified by subsequent heating or chemical alterations in order to be able to faithfully record ancient geomagnetic field through time. Intraplate magmatic events are therefore an important target as they can be well-preserved within continental interiors at distance from mountain-building events along continental margins. However, magmatic events such as the formation of large igneous provinces that are often associated with intraplate magmatism are typically of geologically short duration with the bulk of magmatic products emplaced within 1 Myr. The Midcontinent Rift is an exception as magmatism lasted $\sim$25 Myr from 1109 Ma to 1084 Ma. Additionally, extension ceased in the Midcontinent Rift prior to lithosphere separation, preserving volcanics, intrusives, and sediments of the rift within the continental interior. As a result, rocks of the rift have simple paleomagnetic behavior and form a central record of Mesoproterozoic paleogeography \cite{Swanson-Hysell2021c}. The simple magnetizations held by igneous rocks that formed through an extended duration from 1109 to 1084 Ma that is longer than the Neogene Period have the potential for a well-preserved window into the intensity of Earth's magnetic field .

Despite the exquisite preservation of the rocks, non-ideal paleointensity behaviors and overall low success rates have often led to high failure rates and challenged the interpretation of many previous paleointensity results from the Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a}. One typical failure of paleointensity experiment is the double-slope behavior, as two distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope: a higher paleointensity estimate from the low-temperature portion and a lower paleointensity estimate from the high-temperature portion. Ref. \citealp{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas ref. \citealp{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Results with such non-ideal behavior were rejected by ref. \citealp{Sprain2018a} who applied stricter paleointensity selection criteria.

In this study, we target a unique rock type - high-purity anorthosite xenoliths. These anorthosite xenoliths were brought to the near surface with magma that formed sills of the Beaver River diabase \cite{Zhang2021b}. They cooled and acquired their magnetizations in conjunction with the host diabase. Paleointensity experiments on the anorthosite xenoliths have a very high success rate, yielding consistent specimen- and site-level paleointensity results. Rock magnetic data reveal that the anorthosite xenoliths have low anisotropy of thermal remanent magnetization (TRM) and can acquire TRM linearly. Magnetic imaging shows the anorthosite specimens have dominant magnetic carriers interstitial to and within plagioclase crystals and they do not display strong preferred orientations. Our single-slope, high quality paleointensity data add to the reliable paleointensity records during the $\sim$25 Myr of Midcontinent Rift magmatic activity. Our new data from the Midcontinent Rift intrusive rocks support previous data from the volcanics that there was a strong late Mesoproterozoic geodynamo.

@@

While magmatic activity within the Midcontinent Rift was protracted, there were volcanism and intrusions characterized by particularly rapid and voluminous emplacement. The \textit{ca.} 1092 Ma Beaver Bay Complex in northern Minnesota punctuates one such period of magmatism during the Midcontinent Rift main stage activity. The magma that formed the 1091.7 $\pm$ 0.2 Ma Beaver River diabase dikes and sills of the Beaver Bay Complex transported numerous anorthosite xenoliths that have short-axis diameters up to 180 meters via wide conduits \cite{Boerboom2004a, Boerboom2006b}. The anorthosite xenoliths are plagioclase cumulates that formed comagmatically with the host diabase in the lower crust - an interpretation confirmed by U-Pb zircon geochronology \cite{Zhang2021b}.

@@

Thermal modeling results and paleomagnetic directional data show the anorthosite xenoliths to have acquired thermal remanent magnetizations while cooling with the Beaver River diabase \cite{Zhang2021b}. Step-wise thermal demagnetizations show the anorthosite xenoliths have dominantly single-component magnetization that often unblock sharply within temperature ranges between 500\textdegree C and 580\textdegree C \cite{Zhang2021b}.

@@

Petrographic imaging shows the dominantly monomineralic anorthosite xenoliths of the Beaver River diabase often develop granoblastic texture characterized by equigranular crystals with poor development of petrofabrics (Fig. \ref{fig:Petro_QDM}A,C). In addition, the plagioclase crystals of the anorthosite xenoliths are often free of large inclusions. Magnetic imaging using the quantum diamond microscope (QDM) shows that the dominant magnetic remanence carriers often exist within and sometimes interstitial to plagioclase crystals (Fig. \ref{fig:Petro_QDM}E).

These features of the Beaver River anorthosite xenoliths are distinct from other anorthosites such as the anorthosite xenoliths within the Duluth Complex Anorthositic Series rocks. The plagioclase of the Duluth Complex anorthosites typically develop interlocking textures that display strong igneous foliations (Fig. \ref{fig:Petro_QDM}B) and associated titanomagnetite mineral fabrics characterized by having aligned needles within plagioclase crystals (Fig. \ref{fig:Petro_QDM}D). Magnetic imaging confirms that the titanomagnetite needles within the plagioclase crystals are often strongly magnetic with magnetic moments of individual needles oriented along their long axes (Fig. \ref{fig:Petro_QDM}D,F). Titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex plagioclase \cite{Miller1990a}.

The lack of symplecites and exsolved titanomagnetite needles within the Beaver River anorthosite xenoliths distinguish them from plagioclase cumulates of the Duluth Complex and other layered mafic complex where anisotropic magnetic mineral fabrics associated with igneous foliation often occur \cite{Scofield1986a, Selkin2000a, Feinberg2006a}.

@@

Following IZZI-modified paleointensity experiments \cite{Yu2004a}, 40 from a total of 86 anorthosite specimens and 0 out of a total of 69 diabase specimens passed our paleointensity result selection criteria of having (1) a maximum angular deviation (MAD; ref. \citealp{Kirschvink1980a}) of $<$20\textdegree; (2) scatter parameter ($\beta$; ref. \citealp{Coe1978a}) values of $<$15 per cent; (3) a deviation angle (DANG; ref. \citealp{Tauxe2004a}) of $<$5\textdegree; (4) fraction of remanence fitted for paleointensity estimate (FRAC; ref. \cite{Shaar2013a}) $>$0.6; (5) scatter statistic (SCAT; ref. \citealp{Shaar2013a}) = TRUE; (6) a maximum magnetic moment difference between adjacent zero-field steps (GAP-Max; ref. \citealp{Shaar2013a}) $<$ 0.25; (7) number of pTRM checks $>$ 2; (8) and number of measurements used for paleointensity determination $\geq$ 4. 7 anorthosite sites and no diabase site-level results pass our selections. Example NRM/TRM plots are shown in Fig. \ref{fig:IZZI_examples}. Summary specimen absolute paleointensity estimates and site-level weighted mean paleointensity values are plotted in Fig. \ref{fig:PINT_cooling_corrected} (and provided in Table S1). The cooling rate-corrected absolute paleointensity estimates from all anorthosite sites have a weighted mean of 52.72 $\pm $5.72 $\mu$T. This corresponds to a calculated virtual dipole moment of $\sim$89 ZAm$^2$ ($10^{21} Am^2$) \textit{ca.} 1092 Ma. All measurement-level paleointensity experiment data is available within the MagIC database (\url{}; UPDATE WHEN DOI IS GENERATED).

@@

Paleointensity experiment results from the anorthosite xenoliths and their host diabase show three general types of dominant specimen behaviors on NRM/TRM plots (Fig. \ref{fig:IZZI_examples}): dominantly single-slope, double-slope or sagging, and zigzagging. Typical specimen paleointensity behavior of the anorthosites display straight, single-slope NRM/TRM plots and the accepted fractions of temperature steps span over the origin-trending, primary remanence components (Fig. \ref{fig:IZZI_examples}). We accept all specimen- and site-level absolute paleointensity results from the anorthosite xenoliths that pass the selection. Diabase specimens failed the selection because of double-slope or sagging behavior (fail FRAC selection; see methods), poor pTRM checks, and zigzagging plots (fail SCAT, DANG selection; see methods; Fig. \ref{fig:IZZI_examples}). A 20 mT alternating field treatment after in-field heating steps was applied to some specimens but it did not result in significant changes in paleointensity behaviors of the anorthosite xenoliths or the diabase (Fig. \ref{fig:PINT_cooling_corrected}; Supporting Information).

In addition to estimating paleointensities by introducing a set of selection criteria to filter our experiment results, we apply an independent method of ref. \citealp{Cych2021a} which uses all experimental data regardless of their NRM/TRM plot statistics to perform bias corrected estimation of paleointensity (BiCEP). This Bayesian probabilistic method is based on an assumption that paleointensity estimates from specimens that come from a same cooling unit are distributed around a true paleointensity value with the various deflections being expressed as the curvature parameter of the NRM/TRM plot \cite{Arai1963a, Paterson2011a}. The posterior paleointensity distributions for sites with high-quality specimen-level data (Fig. \ref{fig:PINT_cooling_corrected}) are in agreement with site-level weighted-mean results (Fig. S3).

Overall, the high-quality experiment results from the anorthosite xenoliths of the Beaver River diabase and the consistent paleointensity estimates based on independent methods confirm that the anorthosite xenoliths record a high geomagnetic field \textit{ca.} 1092 Ma. The paleointensity quality index (Q$_{PI}$; ref. \citealp{Biggin2014a}) for the anorthosite xenoliths are all above 4 (Table S2).

@@

Rock magnetic data support that the anorthosite specimens that pass the paleointensity selection criteria have dominant magnetic remanence carriers with magnetic properties similar to stochiometric, non-interacting, single domain magnetite, whereas anorthosite samples that failed the paleointensity result selection and all diabase samples have more pronounced populations of non-ideal carriers. Magnetic property measurement system (MPMS) data show both diabase and anorthosite have dominant magnetic mineralogy of magnetite as evidenced through the Verwey transition (Fig. S1; ref. \citealp{Verwey1939a}). More, anorthosite chips from sites that yield successful paleointensity results have Verwey transition temperatures near 120 Kelvin (K) as expected for stochiometric low-Ti magnetite \cite{Ozdemir1993a}. However, diabase and anorthosite specimens that did not pass our paleointensity selection typically have suppressed Verwey transition toward lower temperatures (Fig. S1), indicating that the specimens likely have a magnetic mineralogy with variable degrees of oxidation or relatively high Ti content \cite{Ozdemir1993a}.

Coercivity spectra from backfield experiments further support the interpretation that rocks that pass or fail the selection criteria have distinct magnetic mineralogy. A compilation of median destructive field (MDF) in Fig. \ref{fig:coercivity} shows that specimens from anorthosites that pass paleointensity selections have a distinctly higher average MDF range ($\sim$60 mT) than other anorthosites and diabase ($\sim$30 mT). Single-component fits for coercivity spectra \cite{Maxbauer2016a} show that anorthosites that yielded successful paleointensity results can have magnetic grain populations with peak coercivity around 80 mT (Fig. \ref{fig:coercivity}). In contrast, other anorthosite and diabase tend to have lower coercivities ($\sim$30 mT). This is consistent with the interpretation that they lack a dominance of single domain magnetite grains in their magnetic mineral assemblages and thus have non-ideal paleointensity behaviors.

\begin{figure}
\noindent\includegraphics[width=\linewidth]{coercivity.pdf}
\centering
\caption{\footnotesize{Top: Example coercivity unmixing plots of anorthosite and diabase specimens from sites that pass or fail our paleointensity selection criteria. Bottom: Box plots of median destructive field (MDF) values for all anorthosite and diabase specimens with single-component coercivity unmixing results. Both plots show that anorthosite specimens that pass paleointensity selection criteria have distinctly higher portion of single-domain-like magnetic mineralogy than other anorthosite and diabase.}}
\label{fig:coercivity}
\end{figure}

\subsection*{TRM anisotropy and nonlinear acquisition check}
Significant remanence anisotropy has been documented to exist within anorthositic rocks that form in layered intrusive complexes \cite{Selkin2000a, Feinberg2006a}. Ref. \citealp{Selkin2000a} demonstrated that strong remanence anisotropy associated with the igneous foliations developed within an anorthosite sample from the Stillwater Complex can lead to significant overestimate or underestimate of paleointensity values depending on the relative orientations between the external applied field magnetic anisotropy axes of a specimen. To assess whether our paleointensity estimates are biased by remanence anisotropy, we calculated the gamma statistic, which is the angular difference between the last pTRM step of paleointensity experiment and the applied field direction. The results show that the anorthosite specimens used for paleointensity experiment have gamma values ranging from $\sim$0.9\textdegree to $\sim$11.9\textdegree, with a median value of 4.2\textdegree (Table S1). These gamma values are similar to those of the Midcontinent Rift volcanics \cite{Sprain2018a}. Therefore, the Beaver River anorthosite xenolith bulk samples do not have a significant remanence anisotropy. Paleodirectional data from our anorthosite xenoliths further support that they have minimal remanence anisotropy as their site mean directions closely match those of the Beaver River diabase hosts \cite{Zhang2021b}. In addition, no strong magnetic fabric was observed in the anorthosite xenoliths through magnetic imaging (Fig. \ref{fig:Petro_QDM}).

We applied full TRMs to a set of our anorthosite specimens that pass the paleointensity selection criteria with lab fields of 30, 50, 70, and 90 $\mu$T along specimen vertical axes. The results show that our anorthosite specimens did not saturate upon the applied fields and there is thus no need for correcting nonlinear thermal remanence acquisitions (Fig. S4).

@@

\subsection*{Cooling rate correction}
Paleointensity estimates from igneous intrusions can also be biased toward higher values due to slow cooling rates \cite{Halgedahl1980a}. It has been shown that large differences in cooling rates between acquisition of an NRM versus a laboratory pTRM can result in overestimated paleointensities for SD grains \cite{Dodson1980a, Halgedahl1980a, Nagy2021a}. From the thermal history model of ref. \citealp{Zhang2021b} we can estimate the duration which the diabase and anorthosite cooled from the Curie temperature of magnetite ($\sim$580\textdegree C) to the time when they fully block in their characteristic natural remanence magnetization ($\sim$500\textdegree C; ref. \citealp{Zhang2021b}). We find the cooling time to be on the order of $\sim$1 kyr, which corresponds to a cooling rate of $\sim 1.6\times10^{-9}$ $^\circ$C s$^{-1}$. In contrast, the lab cooling rate is much faster, at $\sim 1.3\times10^{-1}$ $^\circ$C s$^{-1}$, assuming that it takes 1000s in lab condition to cool through the same temperature interval. This significant cooling rate difference leads to a predicted $\sim$36\% overestimate of true ancient field following the model of ref. \citealp{Halgedahl1980a} (Fig. S2). This estimate on cooling rate effect is similar to that derived from the model of ref. \citealp{Nagy2021a}. We therefore correct our paleointensity results by a factor of 0.74. The cooling rate-corrected specimen paleointensity estimates together with specimen- and site-level weighted means are shown in Fig. \ref{fig:PINT_cooling_corrected}.

\section*{Proterozoic geodynamo}
To best characterize the geomagnetic axial dipole field intensity during a certain time period requires a paleointensity dataset to cover a sufficient amount of time such that paleosecular variations of the geomagnetic field are averaged and the geocentric axial dipole hypothesis can be assumed. Thermal modeling results from ref. \citealp{Zhang2021b} suggest that the Beaver River anorthosite xenoliths were likely heated to tholeiitic magma temperature and acquired thermal remanent magnetization during cooling with their diabase host on a time scale of a few thousand years. In addition, anorthosite xenoliths have paleointensity results that are consistent within small regions but vary between localities. Anorthosite AX12 and AX16 have very similar paleointensity estimates that are distinct from those of sites AX11 and AX13 which were emplaced $\sim$10 km away (Fig. \ref{fig:Geologic_map}). This is consistent with the interpretation that differential cooling history due to heterogeneity of the timing and depths of the Beaver River diabase intrusions caused the anorthosites to capture paleosecular variations \textit{ca.} 1092 Ma. 
% Overall, none of the anorthosite xenoliths record a site mean VDM value lower than 40 ZAm$^2$ (Fig. \ref{fig:PINT_compilation}).
